Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-12-05 paper.tex
Update Table II HMNS luminosities for new compute-eta method
Update HMNS text to describe new compute-eta data in Table II

We use a single time snapshot from that configuration
at $t=11\,{\rm ms}$ after merger.
In this way we approximate the system as stationary over a light-crossing time
of around 1~ms, which is far below the thermal timescale of the remnant.
The volume data are represented on a pseudospectral domain of a
sphere with concentric shells extending to $r\approx1400$~km.
Radial grid spacings are $\Delta r \approx 0.15$~km in the star
and $\Delta r \approx 2.5$~km in the disk,
with 12 cells spanning polar angles $\theta\in[0,\pi]$
and 24 cells spanning azimuthal angles $\phi\in[0,2\pi]$.
\footnote{TODO: way too-low angular resolution}
Though a pseudospectral representation of non-smooth hydrodynamic data
introduces some Gibbs-like oscillations in the variables,
we choose to use this representation instead of the mesh-refined
finite difference grid of the evolution because the pseudospectral
representation uses much less memory:
95~MB in pseuspectral vs.\ 3~GB in finite difference representation.
Obviously, memory loads of this order are not insurmountable;
but they would require some modifications to our
volume data interpolation infrastructure.
For the purposes of this analysis the pseudospectral representation is adequate.
Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat}
show slices of density and temperature from the finite difference volume data.

@@

Each observer samples the distribution function over a uniform grid
in energy, cosine of polar angle, and azimuthal angle with extents
$\{N_{\tilde{\varepsilon}},N_A, N_B\}$ spanning the ranges
$\tilde{\varepsilon}\in(0,100)$~MeV,
$\cos A\in\left((\cos A)_{\rm min}, 1\right)$, and
$B\in[0,2\pi)$.
The integrals for fluxes of number and energy
(Eqns.~\ref{eqn:Ha}, \ref{eqn:Ka}, and \ref{eqn:J_H_S_eps_integrated})
become sums over all rays.
Taking these fluxes in the radial direction,
and using the midpoint rule to convert the integral to a sum, we have
\begin{align}
  \label{eqn:Kr_minkowski}
  \tilde{K}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^2 \cos A_m f_m,\\
  \label{eqn:Hr_minkowski}
  \tilde{H}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^3 \cos A_m f_m,
\end{align}
with $f_m \equiv f(\tilde{\varepsilon}_m,\cos A_m, B_m)$,
$\Delta \equiv \Delta\tilde{\varepsilon}\, \Delta(\cos A)\, \Delta B$,
$N \equiv N_{\tilde{\varepsilon}}\, N_A\, N_B$,
and $m$ the index labeling each ray.
Average energies are given by
$\langle\tilde{\varepsilon}\rangle(\theta)=H^r(\theta)/K^r(\theta)$.
We choose $N_{\tilde{\varepsilon}}=30$, $N_A=120$, and $N_B=20$.
For the \emph{scat} case we choose $(\cos A)_{\rm min}=0.81$;
For the \emph{noscat} case, presenting a smaller emitting region,
we choose $(\cos A)_{\rm min}=0.91$.

We also combine measurements from all observers to estimate
total luminosities and averages over the sky.
We assume the disk is azimuthally symmetric and
symmetric with respect to reflection across the equatorial plane.
Figs.~\ref{fig:nsns_rho_merid} and \ref{fig:nsns_rho_equat} show how these
assumptions are approximately valid at 11~ms after merger.
Total luminosities are then computed as integrals of the radial fluxes
over $\cos\theta$.
Using the trapezoid rule the number and energy luminosities become sums:
\begin{align}
  \label{eqn:luminosity_R}
  R &= C
  \left(\frac{1}{2}(\tilde{K}^r_0+\tilde{K}^r_{N_\theta-1})
  +\sum\limits_{m=1}^{N_\theta-2}\tilde{K}^r_m\right),\\
  \label{eqn:luminosity_L}
  L &= C
  \left(\frac{1}{2}(\tilde{H}^r_0+\tilde{H}^r_{N_\theta-1})
  +\sum\limits_{m=1}^{N_\theta-2}\tilde{H}^r_m\right),
\end{align}
with $C \equiv 4\pi r^2\, \Delta(\cos \theta)$,
\footnote{TODO: source of GR error in coordinate $r$}
$\tilde{K}^r_m \equiv \tilde{K}^r(\theta_m)$, and
$\tilde{H}^r_m \equiv \tilde{H}^r(\theta_m)$.
Average energies over the whole sky are then
$\langle\tilde{\varepsilon}\rangle=L/R$.

\begin{table}%[h]
  % Total fluxes over the sky integrated in google sheet: 20170726-theta_distribution_integrations
  % RTDATA_NO=$DIR/PositionDistributions-nu?-250km-Lev1-e30a120b20-noscat-01.dat
  % RTDATA_GR=$DIR/PositionDistributions-nu?-250km-Lev1-e30a60b10-scat-01.dat
  % DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170324-ray_tracing_nsns/Run03/PositionDistributions
  \caption{
    Comparison of total luminosities and average energies
    of the hypermassive neutron star configuration
    (presented in Sec.~\ref{ssec:test_disk_comparison}).
    The methods are ray tracing using the \emph{noscat} method `RT noscat',
    ray tracing using the \emph{gray} \emph{scat} method `RT scat',
    and the M1 transport simulation `M1'.
    The ray tracing totals were computed from sums over
    observers placed in the $y$-$z$ plane.
    M1 values are taken from \cite[Figs. 7, 9, 10]{fouc2016-m1_evolve_n}.
    Energy luminosities have units $10^{52}\,{\rm erg}\,{\rm s}^{-1}$,
    number luminosities $10^{57}\,{\rm s}^{-1}$, and
    average energies MeV.
    The $\nu_x$ luminosities are per-species:
    multiply by four to get the total heavy-lepton neutrino luminosities.
  }
  \label{tab:nsns_rt_vs_m1}
  \begin{tabularx}{\columnwidth}{X r r r}
    & {\bf RT noscat} & \,\,{\bf RT scat} & \,\,{\bf M1} \\
    \hline
    $L_{\nu_e}$                                 & 5.62 & 6.05 & 5 \\
    $L_{\bar{\nu}_e}$                           & 10.6 & 11.8 & 12 \\
    $L_{\nu_x}$                                 & 23.3 & 12.7 & 12 \\
    $R_{\nu_e}-R_{\bar{\nu}_e}$                 & -1.22 & -1.85 & -2 \\
    $\langle \varepsilon_{\nu_e} \rangle$       & 12.08 & 11.91 & 12 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle$ & 16.07 & 14.67 & 15 \\
    $\langle \varepsilon_{\nu_x} \rangle$       & 34.09 & 23.58 & 26 \\
    \hline
  \end{tabularx}
\end{table}

Tab.~\ref{tab:nsns_rt_vs_m1} compares the all-sky
\footnote{TODO: get accurate M1 numbers from Francois}
luminosities and average energies from ray tracing and from
the M1 transport simulation,
which serves as our fiducial calculation.
Uncertainties are around 15\% for energy luminosities and
10\% for average energies from the M1 simulation
\cite[Sec.~A.6]{fouc2016-m1_evolve_n}.
Because the model was evolved with a gray M1 transport scheme,
the only \emph{scat} method we can use in our ray tracing is the
\emph{gray} method.
As in the post-bounce configuration in Sec.~\ref{ssec:test_collapse},
the \emph{scat} treatment is more faithful than the \emph{noscat} treatment,
significantly so for the heavy-lepton neutrinos.
(Though, we note the energy luminosity for $\nu_e$ is overestimated
20\% by both \emph{noscat} and \emph{scat} treatments,
and the overestimate is slightly worse for the \emph{scat} treatment.)
\footnote{TODO: explain}
With scattering turned off ray tracing overestimates $L_{\nu_x}$ by 200\%;
but with scattering turned on agreement in all measures
is better than 10\%, except for $L_{\nu_e}$ which is overestimated by 20\%.
As the M1 transport evolution also demonstrates,
$R_{\bar{\nu}_e}$ dominates over $R_{\nu_e}$
\footnote{TODO: Need $R$ magnitude not just difference}
because the disrupted neutron star material is releptonizing.

>>>>

We use a single time snapshot from that configuration
at $t=11\,{\rm ms}$ after merger.
In this way we approximate the system as stationary over a light-crossing time
of around 1~ms, which is far below the thermal timescale of the remnant.
The volume data are represented on a pseudospectral domain of a
sphere with concentric shells extending to $r\approx1400$~km.
Radial grid spacings are $\Delta r \approx 0.15$~km in the star
and $\Delta r \approx 2.5$~km in the disk,
with 12 cells spanning polar angles $\theta\in[0,\pi]$
and 24 cells spanning azimuthal angles $\phi\in[0,2\pi]$.
\footnote{TODO: way too-low angular resolution}
Though a pseudospectral representation of non-smooth hydrodynamic data
introduces some Gibbs-like oscillations in the variables,
we choose to use this representation instead of the mesh-refined
finite difference grid of the evolution because the pseudospectral
representation uses much less memory:
95~MB in pseuspectral vs.\ 3~GB in finite difference representation.
Obviously, memory loads of this order are not insurmountable;
but they would require some modifications to our
volume data interpolation infrastructure.
For the purposes of this analysis the pseudospectral representation is adequate.

Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat}
show slices of density and temperature from the finite difference volume data.
These data were evolved on a rectilinear grid spanning approximately 400~km in
both the $x$ and $y$ directions, and 150~km in the $z$ direction in our
evolution coordinates. We extrapolate the fluid and M1 radiation data from
this grid to the larger ray tracing grid by setting all variables to their
floor values. This simple extrapolation is adequate for ray tracing since
the neutrinos are almost entirely free-streaming outside the smaller domain.

@@

Each observer samples the distribution function over a uniform grid
in energy, cosine of polar angle, and azimuthal angle with extents
$\{N_{\tilde{\varepsilon}},N_A, N_B\}$ spanning the ranges
$\tilde{\varepsilon}\in(0,100)$~MeV,
$\cos A\in\left((\cos A)_{\rm min}, 1\right)$, and
$B\in[0,2\pi)$.
The integrals for fluxes of number and energy
(Eqns.~\ref{eqn:Ha}, \ref{eqn:Ka}, and \ref{eqn:J_H_S_eps_integrated})
become sums over all rays.
Taking these fluxes in the radial direction,
and using the midpoint rule to convert the integral to a sum, we have
\begin{align}
  \label{eqn:Kr_minkowski}
  \tilde{K}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^2 \cos A_m f_m,\\
  \label{eqn:Hr_minkowski}
  \tilde{H}^r &= \Delta
  \sum\limits_{m=0}^{N-1}
  \tilde{\varepsilon}_m^3 \cos A_m f_m,
\end{align}
with $f_m \equiv f(\tilde{\varepsilon}_m,\cos A_m, B_m)$,
$\Delta \equiv \Delta\tilde{\varepsilon}\, \Delta(\cos A)\, \Delta B$,
$N \equiv N_{\tilde{\varepsilon}}\, N_A\, N_B$,
and $m$ the index labeling each ray.
Average energies are given by
$\langle\tilde{\varepsilon}\rangle(\theta)=H^r(\theta)/K^r(\theta)$.
We choose $N_{\tilde{\varepsilon}}=30$, $N_A=150$, and $N_B=20$.
To maintain high angular resolution, we only sample rays that pass within
approximately 120~km of the star's center by setting $(\cos A)_{\rm min}=0.88$.

We also combine measurements from all observers to estimate
total luminosities and averages over the sky.
Since we have chosen an arc of observers isolated to the northern hemisphere and
the $y$-$z$ plane, we may extend these data to the full sky by assuming that
emission is azimuthally symmetric and reflection symmetric across the
equatorial plane. Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat}
indicate the approximate validity of these assumptions at 11~ms after merger.
Total luminosities are then computed as integrals of the radial fluxes
over $\cos\theta$.
Using the trapezoid rule the number and energy luminosities become sums:
\begin{align}
  \label{eqn:luminosity_R}
  R &= C
  \left(\frac{1}{2}(\tilde{K}^r_0+\tilde{K}^r_{N_\theta-1})
  +\sum\limits_{m=1}^{N_\theta-2}\tilde{K}^r_m\right),\\
  \label{eqn:luminosity_L}
  L &= C
  \left(\frac{1}{2}(\tilde{H}^r_0+\tilde{H}^r_{N_\theta-1})
  +\sum\limits_{m=1}^{N_\theta-2}\tilde{H}^r_m\right),
\end{align}
with $C \equiv 4\pi r^2\, \Delta(\cos \theta)$,
\footnote{TODO: source of GR error in coordinate $r$}
$\tilde{K}^r_m \equiv \tilde{K}^r(\theta_m)$, and
$\tilde{H}^r_m \equiv \tilde{H}^r(\theta_m)$.
Average energies over the whole sky are then
$\langle\tilde{\varepsilon}\rangle=L/R$.

\begin{table}%[h]
  % Total fluxes over the sky integrated in google sheet: 20171031-theta_distribution_integrations
  % RTDATA_NO=$DIR/PositionDistributions-nu?-250km-Lev2-e30a150b20-noscat-04.dat
  % RTDATA_SC=$DIR/PositionDistributions-trim-nu?-250km-Lev2-e30a150b20-scat-04.dat
  % DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170324-ray_tracing_nsns/Run03/PositionDistributions
  \caption{
    Comparison of total luminosities and average energies
    of the hypermassive neutron star configuration
    (presented in Sec.~\ref{ssec:test_disk_comparison}).
    The methods are ray tracing using the \emph{noscat} method `RT noscat',
    ray tracing using the \emph{gray} \emph{scat} method `RT scat',
    and the M1 transport simulation `M1'.
    The ray tracing totals were computed from sums over
    observers placed in the $y$-$z$ plane.
    M1 values are taken from \cite[Figs. 7, 9, 10]{fouc2016-m1_evolve_n}.
    Energy luminosities have units $10^{52}\,{\rm erg}\,{\rm s}^{-1}$,
    number luminosities $10^{57}\,{\rm s}^{-1}$, and
    average energies MeV.
    The $\nu_x$ luminosities are per-species:
    multiply by four to get the total heavy-lepton neutrino luminosities.
  }
  \label{tab:nsns_rt_vs_m1}
  \begin{tabularx}{\columnwidth}{X r r r}
    & {\bf RT noscat} & \,\,{\bf RT scat} & \,\,{\bf M1} \\
    \hline
    $L_{\nu_e}$                                 & 5.87 & 5.40 & 5 \\
    $L_{\bar{\nu}_e}$                           & 9.70 & 10.7 & 12 \\
    $L_{\nu_x}$                                 & 23.6 & 12.5 & 12 \\
    $R_{\nu_e}-R_{\bar{\nu}_e}$                 & -0.91 & -1.68 & -2 \\
    $\langle \varepsilon_{\nu_e} \rangle$       & 12.7 & 12.0 & 12 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle$ & 16.0 & 14.8 & 15 \\
    $\langle \varepsilon_{\nu_x} \rangle$       & 34.2 & 23.3 & 26 \\
    \hline
  \end{tabularx}
\end{table}

Tab.~\ref{tab:nsns_rt_vs_m1} compares the all-sky
luminosities and average energies from ray tracing and from
the M1 transport simulation,
which serves for qualitative comparison.
Even if M1 and ray tracing methods both provide faithful measurements
of all-sky luminosities, we do not expect perfect agreement since the two
treatments differ fundamentally. In addition to differences in the physics,
M1 fluxes are integrated over the outer boundary of the finite-difference grid
at radii as small as 75~km,
whereas the ray tracing fluxes are integrated over a sphere at radius 250~km,
introducing a time lag between the two measurements of under a millisecond.
Furthermore uncertainties from the M1 simulation are around
15\% for energy luminosities and
10\% for average energies \cite[Sec.~A.6]{fouc2016-m1_evolve_n}.

Because the model was evolved with a gray M1 transport scheme,
the only \emph{scat} method we can use in our ray tracing is the
\emph{gray} method.
As in the post-bounce configuration in Sec.~\ref{ssec:test_collapse},
the \emph{scat} treatment is more faithful than the \emph{noscat} treatment,
significantly so for the heavy-lepton neutrinos.
For example, with scattering turned off the ray tracing and M1 measurements
of $L_{\nu_x}$ disagree by 200\%;
but with scattering turned on agreement is within 10\%.
And as the ray tracing and M1 measurements both show,
$R_{\bar{\nu}_e}$ dominates over $R_{\nu_e}$ significantly,
because the disrupted neutron star material is releptonizing.
